---
title: 'STAT 6021: Final Project EDA'
author: "Group 1"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

## Libraries

```{r}
library(tidyverse)
library(ggcorrplot)
library(ResourceSelection)
```

## Data

Looking at sleep efficiency, with hours of sleep as the response variable

```{r, echo=FALSE} 

sleep <- read.csv("Sleep_Efficiency.csv")

# Remove ID, bedtime, and wakeup time
sleep <- sleep[, -c(1, 4, 5)]
colnames(sleep)
```

## Cleaning

```{r}
# Turn percentages to proportions
sleep <- sleep %>% 
  mutate(REM.sleep.percentage = REM.sleep.percentage / 100,
         Deep.sleep.percentage = Deep.sleep.percentage / 100,
         Light.sleep.percentage = Light.sleep.percentage / 100)

# Weighted sleep (Sleep.duration * Sleep.efficiency)
sleep <- sleep %>% 
  mutate(weighted.sleep = Sleep.duration * Sleep.efficiency)
```

```{r}
str(sleep)
```


## Summary Statistics

```{r}
#sleep duration
summary(sleep$Sleep.duration)
```


```{r}
#sleep efficiency
summary(sleep$Sleep.efficiency)
```


```{r}
#weighted sleep
summary(sleep$weighted.sleep)
```


```{r}
# Model for sleep duration with consumption predictors
lm.duration <- lm(Sleep.duration ~ Caffeine.consumption + Alcohol.consumption + Smoking.status + Exercise.frequency, data = sleep)
summary(lm.duration)
```

```{r}
# Model for weighted sleep with consumption predictors
lm.weightsleep <- lm(weighted.sleep ~ Caffeine.consumption + Alcohol.consumption + Smoking.status + Exercise.frequency, data = sleep)
summary(lm.weightsleep)
```

### Model Assumptions

```{r}
# Only predictors
sleep2 <- sleep[,c(13,1,9,10,11,12)]

# Drop missing predictors
sleep2 <- na.omit(sleep2)
```


#### Linearity assumption

```{r}
long <- gather(sleep2, key="predictor", value="value", 
               Age, Caffeine.consumption, Alcohol.consumption,
               Smoking.status, Exercise.frequency)

ggplot(long, aes(x=value, y=weighted.sleep, color=predictor)) +
  geom_point() +
  facet_wrap(~predictor, scale="free_x")
```

The linearity assumption isn't totally met, due to the nature of our predictors. Besides age, which doesn't seem to exhibit any sort of linear relationship, our data is broken up into distinct categories so our x-axis wouldn't be continuous. As a result, a linear relationship isn't super clear.


#### Equal variance and Independence
```{r}
model_pred <- mutate(sleep2, predictions=fitted(lm.weightsleep), 
                        resid=residuals(lm.weightsleep))

ggplot(model_pred, aes(x=predictions, y=resid)) +
  geom_point() + 
  geom_hline(yintercept = 0, color="aquamarine4")
```

The equal variance and independence assumptions appear to be met as the residuals appear to be scattered around 0 and there are no apparent clusters

#### Normality assumption
```{r}
ggplot(model_pred, aes(sample=resid)) +
  stat_qq() + stat_qq_line(color="red")
```

Our model fits the normality assumption as it very closely follows the line of normality for the Q-Q plot. This means that our residuals follow a normal distribution and with it meeting all of the assumptions except for potentially the linear assumption. Therefore, we deem this model to be acceptable

### Case A: Unhealthy
```{r}
caseA <- data.frame(Age=23, Caffeine.consumption=200, Alcohol.consumption=5,
                    Smoking.status="Yes", Exercise.frequency=0)

predict(lm.weightsleep, caseA, interval="prediction")
```

### Case B: Health nut
```{r}
caseB <- data.frame(Age=50, Caffeine.consumption=0, Alcohol.consumption=0,
                    Smoking.status="No", Exercise.frequency=5)

predict(lm.weightsleep, caseB, interval="prediction")
```

### Case C: Average College Student on weekend
```{r}
caseC <- data.frame(Age=22, Caffeine.consumption=100, Alcohol.consumption=3,
                    Smoking.status="No", Exercise.frequency=3)

predict(lm.weightsleep, caseC, interval="confidence")
```

